|

### 2024-12-13 Drafts/Paper_1/05-discussion.Rmd
Incremental update
Working through Fall updates in hopes I get done before the winter

Here we deconstructed the ecological demographics of numerous harvested marine fish populations . In general, these populations have latent capacity for growth and, even when accounting for the influence of fishing, these growth rates are very close to 1 in general.  There is evidence that fisheries stocks are fished in such a way that they will tend to decline until they have declined below 30% of the maximum population size observed, after that most fished populations tend to experience population growth rates above 1. Most populations appear to exhibit evidence for negative density dependence, which is in line with general beliefs within fisheries science and population ecology more generally, however there are some populations where the evidence suggests some level of density independence or weak positive density dependence , including most of the *G. morhua* populations in NW Atlantic.  Both variability and median $\lambda_{F=0}$ influence recovery, stocks with a relatively high variability in $\lambda_{F=0}$ had a moderate-high recovery potential even when their median $\lambda_{F=0}$ was below one. Effectively recovery of the stocks was likely within 20 years whenever the median $\lambda_{F=0}$ was greater than 1.06 or the standard deviation of lambda was greater than 0.2.

For every population in this analysis the $\lambda_{F=0}$ at the minimum observed abundance was estimated to be greater than one; suggesting all of the populations have the potential for some degree of recovery given enough time when there is no fishing mortality. However, there are seven populations in which $\lambda_{F=0}$ at 40% of the maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* (3) and *G. chalcogrammus* (2)). For both of these Gadus species, there are examples of populations in which the latent capacity for growth is far more robust.  For example, *G. chalcogrammus* is one of the largest fisheries in the world and the vast majority of the landings are from the Eastern Bering Sea population which has a high latent capacity for growth. However the $\lambda_{F=0}$ estimates for two neighboring populations in Alaska are much lower, and it is likely that these populations experience much lower fishing pressure because of their low latent capacity for growth.  
In contrast, in the NW Atlantic, several *G. morhua* populations supported large distant waters fishery for centuries (citations), however the data used in this analysis indicated the majority of the stocks in the NW Atlantic have minimal recovery potential. The maximum abundance used for these populations was likely a small fraction of the historic maximum abundance of these stocks, primarily due to over fishing by the industrial fishery that developed in the latter half of the 20th century (Cite Rose book maybe?, others). Here it is likely that the fishery induced changes to the population, which combined with changes in the environment (Cite cold water and seals) has resulted in these populations having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod population will go extinct unless natural mortality declines significantly (Doug’s predator pit stuff); the results of this analysis largely align with these suggestions given the low recovery potential of this stock and it's extremely low latent capacity for growth. However, as one moves towards the eastern Atlantic, the low recovery potential of *G. morhua* populations was no longer observed, with stocks around mainland Europe all having a high recovery potential. As with *G. chalcogrammus*, this indicates that regional variability is a factor in the ability of populations to withstand fishing pressure. While some Gadus populations have an ability to withstand large fisheries due to their latent capacity for growth, the life-history at a species level is insufficient to predict how a population will respond to fishing. Moreover, this analysis is relevant for only the period under study, the recovery potential assumes that the underlying population demographic rates in the future will be similar to those observed in the past. However, as is likely the case with the *G. morhua* stocks in the NW Atlantic, the response for a population may change over time due to environmental, ecological, and anthropocentric factors (CITE!).

More generally, we observed that populations with high variability in $\lambda_{F=0}$ have a high recovery potential, however, this variability does not necessarily follow on from the expectations of conventional life-history theory (Cite Kingsvader). For example, *C. harengus* are a small, relatively short lived, low trophic level pelagic species (Cite!), based on life-history theory *C. harengus* should exhibit highly variable population growth potential in which they ‘exploit’ opportunities within the environment (CITE). However, the variability of $\lambda_{F=0}$ for the *C. harengus* populations was much lower than the *G. morhua* and *M. aeglefinus* populations found throughout the NE Atlantic. Other small pelagic populations, such at *A. tobianus* and *T. esmarkii* do have elevated  $\lambda_{F=0}$ as life-history theory would expect for small low trophic level species. However, caution would be advised in suggesting that other populations of these species would have similar population dynamics, especially given these two populations are found in the North Sea, which appears to be an environment in which high variability in $\lambda_{F=0}$ is commonplace. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. As such, for populations in which recovery plans and conservation targets are being developed, the developers must account for *both* the magnitude and the frequency of these events. Populations do not *always* increase at their maximum intrinsic rate of growth when at low abundance, but they may have the ability to do so occasionally.

>>>>

@hobdayEcologicalRiskAssessment2011 is a paper using life history for closely related species.

Exploring demographic variability and realized compensatory responses in marine fish populations highlight several instances where stock dynamics do not conform to expected theoretical relationships, which has potentially serious implications for future management. Although most stocks exhibited a latent capacity for annual growth, and exhibited evidence for negative density dependence, overall theoretical growth rates in the absence of fishing were surprisingly close to 1, with several stocks having no apparent capacity for abundance increase in the majority of years. Quota-based fisheries management strategies often aim for neutral (50%) risk in the probability of population decline when setting harvest advice (ref, ref), so it was not unexpected that realized population growth rates centered around 1.

What was unsettling was how little capacity for increase these stocks seemed to retain when excluding the effects of fishing (i.e., $\lambda_{F=0}$). If fishing ceased, future recovery was likely within 20 years whenever the median  $\lambda_{F=0}$ was greater than 1.06 or the standard deviation of lambda was greater than 0.2. This applied to only ADD SOMETHING and none of the highest trophic level species. Widespread distributions, high egg production per female, and broadcast spawning strategies have all contributed to the impression that marine fishes should have very high potential for annual population growth [(CITE)]. When environmental variation results in favorable conditions, stocks are anticipated to exhibit high survival, resulting in large recruitment events or higher overall survival [(CITE)]. While we found indirect evidence of this, in that populations with high variability in  $\lambda_{F=0}$ also had high recovery potential, the magnitude of intra-species variability did not conform to expectations from conventional life-history theory [@kindsvaterTenPrinciplesEvolutionary2016]. For example, Atlantic herring are a small, relatively short lived, low trophic level pelagic species [CITE]. Their life history characteristics are considered to result in highly variable population growth potential which should allow them to exploit favorable environmental conditions and increase quickly in abundance [CITE]. However, the variability of  $\lambda_{F=0}$ for Atlantic herring was much lower than for the higher trophic level and longer lived Atlantic cod and haddock stocks in the same region. Furthermore, these three species co-occur in the NE Atlantic [CITE] and the timespan of data collection was similar, suggesting they likely experienced similar variability in oceanographic conditions. Other small pelagic populations, such as $A. tobianus$ and $T. esmarkii$ were characterized by higher  $\lambda_{F=0}$ values, which conforms to theoretical expectations for small, low trophic level species. However, these two stocks were found in the North Sea, which appears to be an environment in which high variability in  $\lambda_{F=0}$ is commonplace for harvested stocks. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. Contrary to assumptions from typical stock-recruit relationships [CITE], stocks do not necessarily exhibit increased compensatory capacity when at low abundance. These data suggest that while most stocks have the capacity for rapid abundance increase, they only do so occasionally even when at low abundance. Thus, accounting for both the magnitude and frequency of years with high population growth become key components during the development of fisheries advice, as well as for the development of recovery plans and conservation targets [CITE ref new Fisheries act stuff? Precautionary approach stuff?].

NOTE THAT THE BECAUSE WE ASSUMED FISHERY REMOVALS WERE CORRECT COMBINED WITH THE PRODUCITIVTY PARAMETERS HAVING RESULTED IN LAMBDA VALUES THAT ARE 5% LOWER, IF THAT DECLINE IN LAMBDA WAS DUE TO FISHING, OUR PROJECTIONS, GENERATION TIME, AND R0 VALUES WOULD ALL BE SOMEWHAT CONSERVATIVE AS TO GET THAT LOWER LAMBDA VALUE WE ATTRIBUTED THIS TO THE PRODUCTIVITY OF THE SYSTEM BEING LOWER, WHEREAS ATTRIBUTING THAT DIFFERENCE SOLELY TO FISHING WOULD HAVE HAD NO IMPACT ON THE NATURAL MORTALITY OR FECUNDITY

Our best understanding of productivity in these assessed fish stocks suggests that equilibrium theory and general life history characteristics (e.g. longevity, trophic level) are poor predictors of population growth rates and recovery potential. For stocks that have collapsed or are considered to be at very low abundance, it would be prudent to evaluate lifetime reproductive rates prior to developing harvest advice. Numerous stocks in our analysis had no apparent capacity to increase in abundance even in the absence of fishing, which would assure further declines if harvesting continued. For example, there were seven populations in which $\lambda_{F=0}$ at 40% of maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* [3] and *G. chalcogrammus* [2]). These stocks exhibited characteristics much more similar to endangered populations [e.g., several Atlantic salmon Designatable Units in Canada, CITE COSEWIC REPORT], which can also have lifetime reproductive rates < 1 [Gibson refs, CITE], in addition, there was little evidence of negative density dependence for these stocks. In our analysis, the most marked example was Atlantic cod in the NW Atlantic. We have little understanding of how large these Atlantic cod stocks would have been originally. The maximum abundance and thus the magnitude of decline used for our analyses represents the outcome of > 100 years of industrial fishing [Cite Rose book maybe?, others]. Similar to salmon, extreme abundance decline combined with changes in the environment [Cite cold water and seals] may have resulted in these stocks having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod population will go extinct unless natural mortality declines significantly [CITE Doug’s predator pit stuff], a conclusion supported by our analysis. However, the low recovery potential of Atlantic cod was no longer observed as one moves towards the eastern Atlantic, with stocks around mainland Europe all exhibiting high recovery potential. As with $G. chalcogrammus$, this indicates that regional variability must factor strongly into a species’ ability to withstand fishing pressure. Given that variability in population growth potential among stocks could be substantially greater than among species (e.g., compare the variability of $\lambda$ between Atlantic cod and Atlantic herring), life history characteristics and/or trophic level appear to be poor predictors of how marine teleosts will respond to fishing. This problem would compound for less data-rich stocks, in that typical assumptions made for data-poor assessments (give examples of assessment methods) would be more likely to lead to inappropriate harvest advice. That life history characteristics did not link strongly to expected productivity is extremely concerning, given the need for sustainable fisheries management of data-poor stocks [CITE] to mitigate risks of overfishing across the globe [CITE].

Our life-table reconstruction enables the underlying demographic rates to be estimated for populations with sufficient age-structured data, this provides an opportunity of understanding the dynamics of these populations using concepts developed in conservation biology and ecology. In doing so, our optimization procedure assumed that both the estimates of abundance and the fishery removals from the stock assessment are correct. Although these assumptions are unlikely to be fully met [CITE], it is important to remember that harvest advice is being derived for these stocks under the same general assumptions. Thus, our analyses reveal that current assessments can be based on vastly different understandings of productivity and the potential for population growth for the same species, even when incorporating similar elements to describe population dynamics (e.g., equilibrium theory). Our analysis suggests that management advice might be fairly accurate for stocks that aren’t overly variable or those that retain high latent capacity for population increase. Low variability would suggest that the stock is not overly influenced by variation in environmental conditions [or that there was little environmental variability in the ecosystem, CITE], while high latent capacity for population increase could indicate the opposite [CITE]. However, it is not immediately apparent what conditions (environmental, ecological and/or anthropogenic) might be the primary determinants of population growth rates for a given cohort or system. Our findings underscore the need for alternative frameworks in which to derive advice, particularly those that incorporate consideration of interspecies relationships and/or environmental conditions under the umbrella of ecosystem-based fishery management [CITE]. They also underscore the need to collect stock-specific data on marine teleosts, rather than using species-level information from global repositories [e.g., Fishbase, Fishlife, CITE] or assuming values from closely related species to derive harvest advice [CITE].

SOMETHING ABOUT EXPLOITATION HISTORY IF THERE ISN'T ALREADY.
MISSING CAVIATS (e.g. exploitation history, assumptions related to assuming model fit and fishing is right, )

Looking to the future, …..

---- End Heathers edits that need tidied up---

Here we deconstructed the ecological demographics of numerous harvested marine fish populations . In general, these populations have latent capacity for growth and, even when accounting for the influence of fishing, these growth rates are very close to 1 in general.

For every population in this analysis the $\lambda_{F=0}$ at the minimum observed abundance was estimated to be greater than one; suggesting all of the populations have the potential for some degree of recovery given enough time when there is no fishing mortality. However, there are seven populations in which $\lambda_{F=0}$ at 40% of the maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* (3) and *G. chalcogrammus* (2)). For both of these Gadus species, there are examples of populations in which the latent capacity for growth is far more robust.  For example, *G. chalcogrammus* is one of the largest fisheries in the world and the vast majority of the landings are from the Eastern Bering Sea population which has a high latent capacity for growth. However the $\lambda_{F=0}$ estimates for two neighboring populations in Alaska are much lower, and it is likely that these populations experience much lower fishing pressure because of their low latent capacity for growth.

In contrast, in the NW Atlantic, several *G. morhua* populations supported large distant waters fishery for centuries (citations), however the data used in this analysis indicated the majority of the stocks in the NW Atlantic have minimal recovery potential. The maximum abundance used for these populations was likely a small fraction of the historic maximum abundance of these stocks, primarily due to over fishing by the industrial fishery that developed in the latter half of the 20th century (Cite Rose book maybe?, others). Here it is likely that the fishery induced changes to the population, which combined with changes in the environment (Cite cold water and seals) has resulted in these populations having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod population will go extinct unless natural mortality declines significantly (Doug’s predator pit stuff); the results of this analysis largely align with these suggestions given the low recovery potential of this stock and it's extremely low latent capacity for growth. However, as one moves towards the eastern Atlantic, the low recovery potential of *G. morhua* populations was no longer observed, with stocks around mainland Europe all having a high recovery potential. As with *G. chalcogrammus*, this indicates that regional variability is a factor in the ability of populations to withstand fishing pressure. While some Gadus populations have an ability to withstand large fisheries due to their latent capacity for growth, the life-history at a species level is insufficient to predict how a population will respond to fishing. Moreover, this analysis is relevant for only the period under study, the recovery potential assumes that the underlying population demographic rates in the future will be similar to those observed in the past. However, as is likely the case with the *G. morhua* stocks in the NW Atlantic, the response for a population may change over time due to environmental, ecological, and anthropocentric factors (CITE!).

More generally, we observed that populations with high variability in $\lambda_{F=0}$ have a high recovery potential, however, this variability does not necessarily follow on from the expectations of conventional life-history theory [@kindsvaterTenPrinciplesEvolutionary2016]. For example, *C. harengus* are a small, relatively short lived, low trophic level pelagic species (Cite!), based on life-history theory *C. harengus* should exhibit highly variable population growth potential in which they ‘exploit’ opportunities within the environment (CITE). However, the variability of $\lambda_{F=0}$ for the *C. harengus* populations was much lower than the *G. morhua* and *M. aeglefinus* populations found throughout the NE Atlantic. Other small pelagic populations, such at *A. tobianus* and *T. esmarkii* do have elevated  $\lambda_{F=0}$ as life-history theory would expect for small low trophic level species. However, caution would be advised in suggesting that other populations of these species would have similar population dynamics, especially given these two populations are found in the North Sea, which appears to be an environment in which high variability in $\lambda_{F=0}$ is commonplace. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. As such, for populations in which recovery plans and conservation targets are being developed, the developers must account for *both* the magnitude and the frequency of these events. Populations do not *always* increase at their maximum intrinsic rate of growth when at low abundance, but they may have the ability to do so occasionally.

### 2024-12-13 Drafts/Paper_1/08-figures.Rmd
Incremental update
Working through Fall updates in hopes I get done before the winter

\newpage
```{r lambda-sd-fig, echo=F, fig.cap="Standard deviation of the $\\lambda$ estimates for each stock, ordered by species, in this analysis. "}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_sd_",what.2.tune,".png"))
```

>>>>

\newpage
```{r lambda-sd-fig, echo=F, fig.cap="The average standard deviation of $\\lambda_{F = 0}$  for Atlantic herring (*C. harenugs*), Atlantic cod (*G. morhua*), haddock (*M. aeglefinus*), and pollock (*P. virens) stocks in the western Atlantic (NAFO regions 2-5), Northern Altantic (Iceland, Faroe's, and the northeast Arctic), and Eastern Atlantic (Irish, Celtic, North, and Baltic Seas). The errorbars represent one standard deviation. "}
knitr::include_graphics(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_sd_",what.2.tune,".png"))
```
